MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-05 main/MSstatsPTM_Main.tex
Notes

\section{Experimental Procedure}

@@

\subsection*{SpikeIn benchmark - KGG enriched}

We evaluated our approach using a custom designed spike-in benchmark experiment where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The four mixtures were compared with known fold changes for the spike-in peptides.

@@

Further details on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

@@

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.

@@

\bf{Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html

>>>>

\section{Experimental Procedures}

@@

\subsection*{SpikeIn benchmark - KGG enriched \todo{label-free?}}

We evaluated our approach using a custom designed spike-in benchmark experiment with known ground truth, where 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance changes of these 20 proteins were the target of the benchmark.  Unmodified peptides from Human Lysate were used as the estimate of global protein abundance changes. All comparisons with respect to human lysate serve as a null model, there was no change in global protein nor PTM abundance between conditions. Additionally, E coli Lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The spike-in peptides were mixed with human lysate to create four mixture conditions. Two sets of data were acquired for each mixture: KGG enriched + LC-MS, and LC-MS only. The four mixtures were compared with known fold changes for the spike-in peptides.

@@

Further details \todo{clarify which details}  on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \todo{weird reference?} \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

\todo{In the caption of each figure, could you clarify which dataset the figure refers to?}

@@

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.  \todo{Maybe clarify what kind of results}

@@

\todo{This belongs to the 'implementation' section}

{ \bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html
